Using structural and evolutionary information to detect and correct pyrosequencing errors in non-coding RNAs

### 2014-02-04 JCB/conclusion_JCB.tex
JCBFinal

%\TODOJerome{Etoffer la conclusion}
In this article we presented a new and efficient way of exploring the mutational landscape of an RNA under structural constraints,
and apply our techniques to identify and fix sequencing errors. In addition, we introduced a new scoring scheme to measure the
likelihood of sequencing errors that combines the classical nearest-neighbour energy model parameters \cite{Turner2010} to the
recently introduced isostericity matrices \cite{Stombaugh2009}. The latter accounting for geometrical discrepancies occurring
during base pair replacements.

@@

We validated our model on the 5s rRNA and 16s rRNA (See Sec.~\ref{sec:results}) and showed that our technology enables us
to recover mutational errors with high accuracy (Area under the ROC curve between $0.95$ and $0.99$ when the mutations are
located in in base paired regions). Interestingly, we observed that using the isostericity matrices alone yields higher performance
than with the nearest-neighbour energy model alone. This finding supports our hypothesis that isostericity matrices provide a
valuable source of informations that can be efficiently used for RNA sequence and structure analysis. Nonetheless, we also found
that the nearest-neighbour model seems to provide a valuable signal when we under-estimate of the number of errors in the input
sequences.

Our techniques are designed to correcting point-wise errors in structured regions (i.e. base paired nucleotides). However,
our software \texttt{RNApyro} can be easily combined with other methodologies previously developed for correcting other types
of sequencing errors such as indels in unstructured regions or repeats \cite{Quinlan2008,Quince:2009uq}. In further works, we
also intend to include in our model errors stemming from insertions or deletions. It is indeed theoretically possible to considers
these scenario within our dynamic programming scheme \cite{Waldispuhl:2002fk} with minor impacts of the complexity.

Finally, we hope that integrating our software to current sequencing pipe-lines used in metagenomics studies will permit to
improve the estimate of microbial diversity.

>>>>

%\TODOJerome{Etoffer la conclusion}
In this article we presented a new and efficient way of exploring the mutational landscape of an RNA under structural constraints,
and apply our techniques to identify and fix sequencing errors. In addition, we introduced a new scoring scheme to measure the
likelihood of sequencing errors that combines the classical nearest-neighbour energy model parameters \citep{Turner2010} to the
recently introduced isostericity matrices \citep{Stombaugh2009}. The latter accounting for geometrical discrepancies occurring
during base pair replacements.

@@

We validated our model on the 5s rRNA and 16s rRNA (See Sec.~\ref{sec:results}) and showed that our technology enables us
to recover mutational errors with high accuracy (Area under the ROC curve between $0.95$ and $0.99$ when restricting predicted mutations to base paired regions). Interestingly, we observed that using the isostericity matrices alone yields higher performance
than with the nearest-neighbour energy model alone. This finding supports our hypothesis that isostericity matrices provide a
valuable source of informations that can be efficiently used for RNA sequence and structure analysis. Nonetheless, we also found
that the nearest-neighbour model seems to provide a valuable signal when the number of errors in the input sequences is under-estimated.

Our techniques are designed to correcting point-wise errors in structured regions (i.e. base paired nucleotides). However,
our software \texttt{RNApyro} can be easily combined with other methodologies previously developed for correcting other types
of sequencing errors such as indels in unstructured regions or repeats \citep{Quinlan2008,Quince:2009uq}.

In furture works, we
also intend to include in our model errors stemming from insertions or deletions. It is indeed theoretically possible to considers
these scenario within our dynamic programming scheme \citep{Waldispuhl:2002fk} with minor impacts of the complexity.
Finally, we hope that integrating our software to current sequencing pipe-lines used in metagenomics studies will permit to
improve the estimate of microbial diversity.

### 2014-02-04 JCB/introduction_JCB.tex
JCBFinal

For half a century, biological molecules have been studied as a proxy to understand evolution~\cite{Zuckerkandl1965}, and due
to their fundamental functions and remarkably conserved structures, rRNAs have always been a prime candidate for phylogenetic
studies~\cite{Olsen1986, Olsen1993}. In recent years, studies as the \emph{Human Microbiome Project}~\cite{Turnbaugh2007} benefited
of new technologies such as the NGS techniques to sequence as many new organisms as possible and extract an unprecedented flow of new information. 
Nonetheless, these high-throughput techniques typically have high error rates that make their applications to metagenomics (a.k.a. environmental
genomics) studies challenging. For instance, pyrosequencing as implemented by Roche's 454 produces may have an error rate raising up to 10\%.
Because there is no cloning step, resequencing to increase accuracy is not possible and it is therefore vital to disentangle noise from true sequence
diversity in this type of data \cite{Quince:2009uq}. Errors can be significantly reduced  when large multiple sequence alignments with close homologs
are available, but in studies of new or not well known organisms, such information is rather sparse. In particular, it is common that there is not enough  similarity to 
differentiate between the sequencing errors and the natural polymorphisms that we want to observe, often leading to artificially inflated diversity estimates~\cite{Kunin2010}.
A few techniques have been developed to remedy to this problem~\cite{Quinlan2008,Medvedev2011} but they do not take
into account all the available information. It is therefore essential to develop methods that can exploit any type of signal available to correct errors.

@@

Here, we expand the range of algorithmic techniques previously introduced with the \RNAmutants software~\cite{Waldispuhl2008,Waldispuhl2011}.
Instead of exploring the full conformational landscape and sample mutants, we develop an inside-outside algorithm that enables us
to explore the complete mutational landscape with a \emph{fixed} secondary structure and to calculate exactly mutational probability values. In addition
to a gain into the numerical precision, this strategy allows us to drastically reduce the computational complexity ($\mathcal{O}(n^3 \cdot M^2)$ for the
original version of  \RNAmutants to $\mathcal{O}(n \cdot M^2)$ for \RNApyro, where $n$ is the size of the sequence and $M$ the maximal number of mutations).

We design a new scoring scheme combining nearest-neighbor models \cite{Turner2010} to isostericity metrics \cite{Stombaugh2009}.
Classical approaches use a Boltzmann distribution whose weights are estimated using a nearest-neighbour energy model~\cite{Turner2010}. However, the
latter only accounts for  canonical and wobble, base pairs. As was shown by Leontis and Westhof~\cite{Leontis2001},
the diversity of base pairs observed in tertiary structures is much larger, albeit their energetic contribution remains unknown. To quantify geometrical discrepancies, 
an isostericity distance has been designed \cite{Stombaugh2009}, increasing as two base pairs geometrically differ from each other in space. Therefore, we 
incorporate these scores in the Boltzmann weights used by \RNApyro.

%\TODOTous{Mentionner technologie Illumina 50-75 nts, et modele d'erreur/mutations.}
We illustrate and benchmark our techniques for point-wise error corrections on the 5S ribosomal RNA. We choose the latter since it has been extensively
used for phylogenetic reconstructions~\cite{Hori1987} and its sequence has been recovered for over 712 species (in the Rfam seed alignment with id
\texttt{RF00001}). Using a leave one out strategy, we perform random distributed mutations on a sequence. While our methodology is restricted to the correction of 
point-wise error in structured regions (i.e. with base pairs), we show that \texttt{RNApyro} can successfully extract a signal that can be used to reconstruct the 
original sequence with an excellent accuracy. This suggests that \RNApyro is a promising algorithm to complement existing tools in the NGS error-correction 
pipeline.

>>>>

For half a century, biological molecules have been studied as a proxy to understand evolution~\citep{Zuckerkandl1965}, and due
to their fundamental functions and remarkably conserved structures, rRNAs have always been a prime candidate for phylogenetic
studies~\citep{Olsen1986, Olsen1993}. In recent years, studies as the \emph{Human Microbiome Project}~\citep{Turnbaugh2007} benefited
of new technologies such as the NGS techniques to sequence as many new organisms as possible and extract an unprecedented flow of new information. 
Nonetheless, these high-throughput techniques typically have high error rates that make their applications to metagenomics (a.k.a. environmental
genomics) studies challenging. For instance, pyrosequencing as implemented by Roche's 454 produces may have an error rate raising up to 10\%.
Because there is no cloning step, resequencing to increase accuracy is not possible and it is therefore vital to disentangle noise from true sequence
diversity in this type of data \citep{Quince:2009uq}. Errors can be significantly reduced  when large multiple sequence alignments with close homologs
are available, but in studies of new or not well known organisms, such information is rather sparse. In particular, it is common that there is not enough  similarity to 
differentiate between the sequencing errors and the natural polymorphisms that we want to observe, often leading to artificially inflated diversity estimates~\citep{Kunin2010}.
A few techniques have been developed to remedy to this problem~\citep{Quinlan2008,Medvedev2011} but they do not take
into account all the available information. It is therefore essential to develop methods that can exploit any type of signal available to correct errors.

@@

Here, we expand the range of algorithmic techniques previously introduced with the \RNAmutants software~\citep{Waldispuhl2008,Waldispuhl2011}.
Instead of exploring the full conformational landscape and sample mutants, we develop an inside-outside algorithm that enables us
to explore the complete mutational landscape with a \emph{fixed} secondary structure and to calculate exactly mutational probability values. In addition
to a gain into the numerical precision, this strategy allows us to drastically reduce the computational complexity ($\mathcal{O}(n^3 \cdot M^2)$ for the
original version of  \RNAmutants to $\mathcal{O}(n \cdot M^2)$ for \RNApyro, where $n$ is the size of the sequence and $M$ the maximal number of mutations).

We design a new scoring scheme combining nearest-neighbor models \citep{Turner2010} to isostericity metrics \citep{Stombaugh2009}.
Classical approaches use a Boltzmann distribution whose weights are estimated using a nearest-neighbour energy model~\citep{Turner2010}. However, the
latter only accounts for  canonical and wobble, base pairs. As was shown by Leontis and Westhof~\citep{Leontis2001},
the diversity of base pairs observed in tertiary structures is much larger, albeit their energetic contribution remains unknown. To quantify geometrical discrepancies, 
an isostericity distance has been designed \citep{Stombaugh2009}, increasing as two base pairs geometrically differ from each other in space. Therefore, we 
incorporate these scores in the Boltzmann weights used by \RNApyro.

%\TODOTous{Mentionner technologie Illumina 50-75 nts, et modele d'erreur/mutations.}
We illustrate and benchmark our techniques for point-wise error corrections on the 5S ribosomal RNA. We choose the latter since it has been extensively
used for phylogenetic reconstructions~\citep{Hori1987} and its sequence has been recovered for over 712 species (in the Rfam seed alignment with id
\texttt{RF00001}). Using a leave one out strategy, we perform random distributed mutations on a sequence. While our methodology is restricted to the correction of 
point-wise error in structured regions (i.e. with base pairs), we show that \texttt{RNApyro} can successfully extract a signal that can be used to reconstruct the 
original sequence with an excellent accuracy. This suggests that \RNApyro is a promising algorithm to complement existing tools in the NGS error-correction 
pipeline.

### 2014-02-04 JCB/methods_JCB.tex
JCBFinal

We introduce a probabilistic model, which aims at capturing both the stability of the folded RNA and its affinity towards a predefined 3D conformation.
To that purpose, a Boltzmann weighted distribution is assumed, based on a pseudo-energy function $\PE{\cdot}$ which includes contributions for both the free-energy and its putative isostericity towards a multiple sequence alignment. In this model, the probability that the nucleotide at a given position needs to be mutated (i.e. corresponds to a sequencing error) can be computed using a variant of the \emph{Inside-Outside algorithm}~\cite{Lari1990} in time which scales linearly with the sequence length.

@@

\subsubsection{Energy Contribution.}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, using values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. neither $\Gb\Ub$, $\Ub\Gb$, $\Cb\Gb$, $\Gb\Cb$, $\Ab\Ub$ nor $\Ub\Ab$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed. For instance, multiloops do not consider base-specific contributions, and therefore their consideration would constitute a criterion for preferring a sequence over another. Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity Contribution.}
The concept of isostericity score is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the isostericity contribution to the pseudo-energy as
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, $0$ corresponding to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution exponentially favors sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.
%\TODOYann{Avoid duplicate reference to Stombaugh et al}

@@

Clearly, the number of sequences in $\mathcal{D}_{s,M}$ grows exponentially with the sequence length, therefore one cannot realistically rely on an exhaustive enumeration to compute the mutational profile. To work around this issue, we propose a linear-time variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990} to compute this probability, based on two sets of dynamic programming equations. 
%from the two  functions $\mathcal{Z}_*^*$ and $\mathcal{Y}_*^*$.

@@

\subsubsection{Inside computation.}
\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{FigDPInsideWrapper}
\caption{Principle of the inside computation (partition function). Any sequence (mutated)  
can be decomposed as a sequence preceded by a, possibly mutated, base 
(Unpaired case), a sequence surrounded by some base-pair (Stacking-pair case), 
or as two sequences segregated by some base-pair (General base-pairing case). In this latter case, mutations must be distributed between sub-sequences.\label{fig:inside}}
\end{figure}

@@

\item {\bf Case 3: Paired leftmost position, but no stacking pairs.} In this case, the structure is split into two parts by the base pair ($\Struct = \op \Struct'\cp\Struct''$). Let us denote by $k$ the partner of position $i$, and by $a$, $b$ and $c$ the bases found at positions $i$, $k$ and $j$ respectively, then one has:
\begin{equation}
\label{eq:Z_rec_C}
	\Z{\Struct}{m}{x,y}=\sum_{\substack{a',b'\in \B^2,\\ \Kron_{ab, a'b'}\le m}}
      \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
   		 e^{\frac{-\EBP{i,k}{\varnothing}{a'b'}}{RT}}
      \cdot\Z{\Struct'}{m-m'-\Kron_{ab,a'b'}}{a',b'}
      \cdot\Z{\Struct''}{m'}{b',y}.
\end{equation}
In other words, if the leftmost position is paired, and the base-pair is not stacking onto another base-pair, then the  only term contributing directly to the energy is the isostericity of the base pair. Admissible sequences for $\Struct$ consist of two paired nucleotides $a'$ and $b'$ at positions $i$ and $k$ respectively, flanking a sequence for $\Struct'$ (over an interval $[i+1,k-1]$), and followed by a (possibly empty) sequence for $\Struct''$ (over $[k+1,j]$). Since the total number of mutations sums to $m$, a parameter $m'$ is introduced to distribute the remaining mutations between the two sequences.
\end{itemize}

@@

\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{FigDPOutsideWrapper}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires the precomputation and storage of the inside contributions.\label{fig:outside}}
\end{figure}
In general, the main recurrence below works by extending the excluded structure $\Struct$ (covering the $]i,j[$ interval) in the leftward direction. As shown in Figure~\ref{fig:outside}, four cases must be considered, depending on the base-pairing status and directionality of the next considered position:
\begin{itemize}
\item {\bf Case 1: Unpaired position.} When the innermost leftward base is unpaired, any base $a\in\B$ may be chosen for this position, and the rest of the sequence is generated recursively. The number of remaining mutations may be decreased by the choice of $a$, leading to the following recurrence:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'\in \B,\\ \Kron_{a,a'}\le m}}
    \Y{\ub\Struct}{m- \Kron_{a,a'}}{a',y}
\label{eq:Y_rec_A}
\end{equation}
\item {\bf Case 2: Stacking base-pair.} If both ends of the excluded structure are paired together, and are nested within another base-pair in the remaining structure, then an additional contribution, stemming from a stacking pairs energy, has to be considered. The outside terms are then computed by simulating any pair of base-pairing nucleotides for $i,j$, and by proceeding recursively on the remaining portion, as follows:
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 e^{\frac{-\EBP{i,j}{xy}{a'b'}}{RT}}\cdot
    \Y{\op\Struct\cp}{m- \Kron_{ab,a'b'}}{a',b'} 
\label{eq:Y_rec_B}
\end{equation}
\item {\bf Case 3: Next position paired rightward, in the absence of stacking pair.} In this case, the innermost leftward position $i$ is paired to the right at some position $k$ . 
Let us assume that its partner resides outside the excluded structure $\Struct$ (This assumption is provably without loss of generality, and directly follows from the fact that $\Struct$ is well-parenthesized). 
The structure within the base-pair may be described by the expression $\op\Struct \Struct'\cp$, where $\Struct'$ may possibly be empty (except if $\Struct$ is enclosed within corresponding brackets, to avoid stacking pairs). Therefore, any sequence considered by the outside computation consists in three independent parts: two nucleotides for the paired positions, a sequence for the region excluding $\Struct''$, and a sequence for $\Struct'$. It follows that:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
  		 e^{\frac{-\EBP{i,k}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct \Struct'\cp}{m- \Kron_{ab,a'b'} - m'}{a',b'}
     \cdot\Z{\Struct'}{m'}{y,b'} .
\label{eq:Y_rec_C}
\end{equation}
\item {\bf Case 4: Next position paired leftward.} If the innermost, leftward, position $i$ is paired to some position $k<i$, delimiting a substructure $\Struct'$, then any sequence considered by the outside computation consists in three parts: two nucleotides $a$ and $b$, a sequence for $\Struct'$, and a sequence for the region excluding $\op\Struct'\cp\Struct$. Consequently, one has:
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
   	 e^{\frac{-\EBP{k,i}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct'\cp\Struct}{m- \Kron_{ab,a'b'} - m'}{a',b}
     \cdot\Z{\Struct'}{m'}{a',b'}
\label{eq:Y_rec_D}
\end{equation}
\end{itemize}

@@

\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{FigDPCombineWrapper}
\caption{Combining the inside and outside contributions to compute the total Boltzmann weight of all sequences having a given base (case 1) or base-pair (cases 2 and 3) at a given position.\label{fig:combine}}
\end{figure}

>>>>

We introduce a probabilistic model, which aims at capturing both the stability of the folded RNA and its affinity towards a predefined 3D conformation.
To that purpose, a Boltzmann weighted distribution is assumed, based on a pseudo-energy function $\PE{\cdot}$ which includes contributions for both the free-energy and its putative isostericity towards a multiple sequence alignment. In this model, the probability that the nucleotide at a given position needs to be mutated (i.e. corresponds to a sequencing error) can be computed using a variant of the \emph{Inside-Outside algorithm}~\citep{Lari1990} in time which scales linearly with the sequence length.

@@

\subsubsection{Energy Contribution.}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, using values from the Turner 2004 model retrieved from the NNDB~\citep{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. neither $\Gb\Ub$, $\Ub\Gb$, $\Cb\Gb$, $\Gb\Cb$, $\Ab\Ub$ nor $\Ub\Ab$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed. For instance, multiloops do not consider base-specific contributions, and therefore their consideration would constitute a criterion for preferring a sequence over another. Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity Contribution.}
The concept of isostericity score is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\citep{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the isostericity contribution to the pseudo-energy as
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\citep{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, $0$ corresponding to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution exponentially favors sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.
%\TODOYann{Avoid duplicate reference to Stombaugh et al}

@@

Clearly, the number of sequences in $\mathcal{D}_{s,M}$ grows exponentially with the sequence length, therefore one cannot realistically rely on an exhaustive enumeration to compute the mutational profile. To work around this issue, we propose a linear-time variant of the
 \emph{Inside-Outside algorithm}~\citep{Lari1990} to compute this probability, based on two sets of dynamic programming equations. 
%from the two  functions $\mathcal{Z}_*^*$ and $\mathcal{Y}_*^*$.

@@

\subsubsection{Inside computation.}
\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{Figure1-FigDPInsideWrapper}
\caption{Principle of the inside computation (partition function). Any sequence (mutated)  
can be decomposed as a sequence preceded by a, possibly mutated, base 
(Unpaired case), a sequence surrounded by some base-pair (Stacking-pair case), 
or as two sequences segregated by some base-pair (General base-pairing case). In this latter case, mutations must be distributed between sub-sequences.\label{fig:inside}}
\end{figure}

@@

\item {\bf Case 3: Paired leftmost position + absence of stacking pairs.} In this case, the structure is split into two parts by the base pair ($\Struct = \op \Struct'\cp\Struct''$). Let us denote by $k$ the partner of position $i$, and by $a$, $b$ and $c$ the bases found at positions $i$, $k$ and $j$ respectively, then one has:
\begin{equation}
\label{eq:Z_rec_C}
	\Z{\Struct}{m}{x,y}=\sum_{\substack{a',b'\in \B^2,\\ \Kron_{ab, a'b'}\le m}}
      \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
   		 e^{\frac{-\EBP{i,k}{\varnothing}{a'b'}}{RT}}
      \cdot\Z{\Struct'}{m-m'-\Kron_{ab,a'b'}}{a',b'}
      \cdot\Z{\Struct''}{m'}{b',y}.
\end{equation}
In other words, if the leftmost position is paired, and the base-pair is not stacking onto another base-pair, then the  only term contributing directly to the energy is the isostericity of the base pair. Admissible sequences for $\Struct$ consist of two paired nucleotides $a'$ and $b'$ at positions $i$ and $k$ respectively, flanking a sequence for $\Struct'$ (over an interval $[i+1,k-1]$), and followed by a (possibly empty) sequence for $\Struct''$ (over $[k+1,j]$). Since the total number of mutations sums to $m$, a parameter $m'$ is introduced to distribute the remaining mutations between the two sequences.
\end{itemize}

@@

\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{Figure2-FigDPOutsideWrapper}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires the precomputation and storage of the inside contributions.\label{fig:outside}}
\end{figure}
In general, the main recurrence below works by extending the excluded structure $\Struct$ (covering the $]i,j[$ interval) in the leftward direction. As shown in Figure~\ref{fig:outside}, four cases must be considered, depending on the base-pairing status and directionality of the next considered position:
\begin{itemize}
\item {\bf Case 1: Unpaired position.} When the innermost leftward base is unpaired, any base $a\in\B$ may be chosen for this position, and the rest of the sequence is generated recursively. The number of remaining mutations may be decreased by the choice of $a$, leading to the following recurrence:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'\in \B,\\ \Kron_{a,a'}\le m}}
    \Y{\ub\Struct}{m- \Kron_{a,a'}}{a',y}
\label{eq:Y_rec_A}
\end{equation}
\item {\bf Case 2: Stacking base-pair.} If both ends of the excluded structure are paired together, and are nested within another base-pair in the remaining structure, then an additional contribution, stemming from a stacking pairs energy, has to be considered. The outside terms are then computed by simulating any pair of base-pairing nucleotides for $i,j$, and by proceeding recursively on the remaining portion, as follows:
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 e^{\frac{-\EBP{i,j}{xy}{a'b'}}{RT}}\cdot
    \Y{\op\Struct\cp}{m- \Kron_{ab,a'b'}}{a',b'} 
\label{eq:Y_rec_B}
\end{equation}
\item {\bf Case 3: Next position paired rightward + absence of stacking pair.} In this case, the innermost leftward position $i$ is paired to the right at some position $k$ . 
Let us assume that its partner resides outside the excluded structure $\Struct$ (This assumption is provably without loss of generality, and directly follows from the fact that $\Struct$ is well-parenthesized). 
The structure within the base-pair may be described by the expression $\op\Struct \Struct'\cp$, where $\Struct'$ may possibly be empty (except if $\Struct$ is enclosed within corresponding brackets, to avoid stacking pairs). Therefore, any sequence considered by the outside computation consists in three independent parts: two nucleotides for the paired positions, a sequence for the region excluding $\Struct''$, and a sequence for $\Struct'$. It follows that:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
  		 e^{\frac{-\EBP{i,k}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct \Struct'\cp}{m- \Kron_{ab,a'b'} - m'}{a',b'}
     \cdot\Z{\Struct'}{m'}{y,b'} .
\label{eq:Y_rec_C}
\end{equation}
\item {\bf Case 4: Next position paired leftward.} If the innermost, leftward, position $i$ is paired to some position $k<i$, delimiting a substructure $\Struct'$, then any sequence considered by the outside computation consists in three parts: two nucleotides $a$ and $b$, a sequence for $\Struct'$, and a sequence for the region excluding $\op\Struct'\cp\Struct$. Consequently, one has:
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
   	 e^{\frac{-\EBP{k,i}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct'\cp\Struct}{m- \Kron_{ab,a'b'} - m'}{a',b}
     \cdot\Z{\Struct'}{m'}{a',b'}
\label{eq:Y_rec_D}
\end{equation}
\end{itemize}

@@

\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{Figure3-FigDPCombineWrapper}
\caption{Combining the inside and outside contributions to compute the total Boltzmann weight of all sequences having a given base (case 1) or base-pair (cases 2 and 3) at a given position.\label{fig:combine}}
\end{figure}
